Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-06-13 papers/paper-eccentricity/paper-eccentricity.tex
- small edits to numerical section

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{rantala2017, rantala2020,mannerkoski2023}, implemented in \gadget{} \citep{springel2021}.
For details about the implementation of \ketju{}, we point the reader to \citet{mannerkoski2023}.

We construct our galaxy models to represent the nuclear bulge of a gas-devoid elliptical galaxy, matching those models of \citet{nasim2020}.
Each galaxy is thus represented as a stellar-only Dehnen sphere \citep{dehnen1993} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed with zero initial velocity.
We test five different mass resolutions by sampling the stellar phase space distribution with a varying number of stellar particles: $N_\star = \{2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6\}$.

We then construct isolated merger initial conditions by placing two galaxies on an elliptical orbit with varying values of eccentricity, at a fixed initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
The chosen eccentricities $e_0$ are 0.90 and 0.99.
We ensure that the radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris2022}.
For each orbital configuration, we run ten realisations for each mass resolution, to account for Brownian motion effects of the SMBH binary caused by the discretised phase space.

@@

\begin{figure}
    \includegraphics{convergence}
    \caption{Convergence of the eccentricity scatter $\sigma_e$ as a function of mass resolution $M_\bullet/m_\star$ and the number of stellar particles per progenitor galaxy $N_\star$ for initial eccentricities $e_0=0.90$ and $e_0=0.99$. 
    The expected $1/\sqrt{N_\star}$ scaling is recovered for the $e_0 = 0.90$ suite after applying a $2.5\sigma$-cut to the simulation data. 
    For the $e_0=0.99$ suite, a steeper scaling than the $1/\sqrt{N_\star}$ is obtained after applying a similar $2.5\sigma$-cut to the data.}
    \label{fig:convergence}
\end{figure}

\subsection{\drafting{Eccentricity scatter in simulations}}

@@

Critically, the eccentricity values beyond $2.5\sigma$ from the mean that are removed in our cleaned samples are each still a valid measurement from the simulation, and highlight the presence of stochasticity in eccentricity for these idealised simulation set-ups with a highly-radial initial orbit. 
The cause of these low eccentricity values is discussed in \autoref{ssec:toy}.

@@

To understand the observed scatter in eccentricity, we investigate the binary binding process in each simulation. 
Whilst the two SMBHs are unbound, they undergo a number of hyperbolic orbits influenced by the galactic merger potential.
At each pericentre passage, the two SMBHs scatter off each other with some deflection angle $\theta_\mathrm{defl}$, defined:
\begin{equation}\label{eq:theta}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, and $L$ is the magnitude of the SMBH 
binary angular momentum vector and $E$ the binary orbital energy at the time of the pericentre passage \citep{binney2008}.
The deflection angle is a function of the impact parameter $b$ of the SMBH trajectory, however is able to be more robustly measured than $b$ from the simulation data.
We set as the first deflection angle the first 
angle $\theta_\mathrm{defl}$ that exceeds an orbit-dependent threshold:
$\theta_{\mathrm{defl,min}}=90\degr$ for the $e_0=0.99$ suite and $\theta_{\mathrm{defl,min}}=45\degr$ for the $e_0=0.90$ suite.
In general, this corresponds to the second or third actual pericentre passage of the SMBHs in the simulations.
An example of $\theta_\mathrm{defl}$ is shown for two realisations of the $e_0=0.90$ ($e_0=0.99$) orbit in the left (right) panel of \autoref{fig:orbit}, which occurs between the points A and B in the SMBH binary orbit in each inset panel. 
We observe evidence for a relationship between the deflection angle $\theta_\mathrm{defl}$ and the resulting eccentricity at the time the SMBH binary becomes hard, which we explain using a simple toy model.

@@

\subsection{Reproducing the eccentricity behavior with a toy model}\label{ssec:toy}

@@

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $12\degr$,
while in the $e_0=0.99$ case a smaller shift of $6\degr$ is used.
These shifts are likely required due to the rotation of the stellar component tilting the background potential
relative to the SMBH trajectories in the simulations, with the effect being stronger in the $e_0=0.90$ case
due to the less radial merger orbit.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta_\mathrm{defl} \gtrsim 120\degr$ region,
although the model does appear to capture the mean behavior fairly well.
The behavior of the model in this region is also quite sensitive to the exact value of $e_\mathrm{s}$,
which might be related to the large scatter in the simulation data.

Otherwise the behavior of the simulation data is captured well,
with the model correctly producing the two main eccentricity minima,
as well as the $e\approx 1$ region between them.
The minima occur for trajectories where the torque from the background potential together with the dynamical friction
causes the SMBHs to loop around into a nearly circular orbit,
while highly eccentric binaries are produced when the binary is trapped into nearly radial oscillations along 
the main axes of the potential, as is seen in \autoref{fig:model_orbits}.

@@

Our results show that the variation in eccentricity $e_\mathrm{h}$ for a given system configuration is tightly correlated with the variation in the deflection angle $\theta_\mathrm{defl}$, independent of the simulation mass resolution.
Together with the simple toy model that reproduces well the behavior seen in the simulations,
this demonstrates that the sensitivity of the SMBH binary eccentricity to slight changes in the merger orbit is a physical feature of galactic mergers on highly radial orbits,
at least in major mergers where the SMBHs have nearly equal masses,
and not an error due to insufficient numerical resolution.

As shown by both the simulations and the toy model, in these mergers the eccentricity of the SMBH binary can span nearly the full possible range between 0 and 1,
depending on comparatively tiny differences on the scale of a parsec in the particular realization of the galactic merger orbit.
\drafting{
    This observation is at odds with \citet{nasim2021}, where binaries that formed at high eccentricity tended to proceed to higher eccentricities.
    However, the tension can be explained if the simulations of \citet{nasim2021} lie in the linear region of the $\theta_\mathrm{defl}$--$e_\mathrm{h}$ parameter space (where most of the simulations in this work lie), coupled with the low-number statistics of \citet{nasim2021}.
    This would suggest that the potential of the merger remnant, as opposed to the rotation of the SMBH binary relative to the stellar background, is the dominant predictor of hard binary eccentricity.
}

Compared to major mergers, in minor mergers where there is less asymmetry in the stellar background when the SMBHs become bound,
the scatter in eccentricity is likely to be low even in the case of a radial merger orbit, since the system is less sensitive to the exact value of the deflection angle,
or the impact parameter, as can be seen from the $e_\mathrm{s}=0.2$ case in \autoref{fig:model_orbits}.

The simulation setup used in this work is highly idealized,
and covers only a small part of the parameter space of merger configurations since we only consider fairly high-eccentricity galactic merger orbits.
However, similar behaviour can be expected to occur also in more realistic models,
since the relevant dynamics occurs within a few hundred parsecs from the centre of the merged galaxy,
where we do not expect any significant  effects from dark matter or the outer parts of the galaxy, which are not included in our simulations.
Lower eccentricity merger orbits can be expected to show less scatter in the binary eccentricity due to the lack of a hard scattering event that is sensitive to slight perturbations in the merger orbit,
and this is also seen in simulation results \citep{gualandris2022}.
However, most major mergers are expected to occur on highly radial orbits based in cosmological simulations \citep[e.g.][]{khochfar2006},
and the dynamical friction from the dark matter halo would generally drive even lower eccentricity merger orbits to highly radial final plunges when the initial separation of the galaxies is large enough.

In numerical simulations with a discretised phase space, the two primary causes of variation in the impact parameter are the initial particle sampling variation in the merger initial conditions and the Brownian motion of the SMBH binary \citep{merritt2001}.
Both of these effects scale as $1/\sqrt{N_\star}$, which can explain the observed scaling of $\sigma_e$ \citep{nasim2020} when the system falls into the region of parameter space where the relation between $e$ and $b$
is approximately linear.
The deviation from this scaling seen in \autoref{fig:convergence}, particularly when not excluding outlying data points, is then simply explained by the fact that the eccentricity is not globally a linear function of the impact parameter.

While the scatter in the impact parameter between different numerical realizations of the same merger is due to relatively low number of particles compared to real galaxies,
we expect that uncertainty of a similar magnitude is also present in real systems due various mechanisms, such as perturbations from their environment.
The SMBHs are likely also not located exactly at the centre of mass of the galaxy,
as e.g.\ work by \citet{batcheldor2010} has suggested that the SMBH in M87 may be displaced from the galactic photo-centre by $\sim7\,\mathrm{pc}$ due to jet acceleration or gravitational recoil kicks. 
It is therefore evident that it is not possible to give an exact prediction for the SMBH binary eccentricity produced by a given merger,
and instead we must focus on predicting the distribution of eccentricities that can be produced by a given merger configuration.
Such predictions are crucial for constraining observational quantities that are correlated with binary eccentricity,
most notably SMBH binary merger timescale and the spectrum of GWs at nanohertz frequencies observable with PTAs \drafting{(refs)}.

@@

P.H.J. acknowledge the support
by the European Research Council via ERC Consolidator Grant KETJU (no. 818930).

>>>>

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{rantala2017, rantala2020,mannerkoski2023}, implemented in \gadget{} \citep{springel2021}.
For details about the implementation of \ketju{}, we point the reader to \citet{mannerkoski2023}.
We use the same code configuration (including post-Newtonian corrections to order 3.5, fast multiple method with second order multipoles, and hierarchical time integration) as \citet{mannerkoski2023}.

We construct our galaxy models to represent the nuclear bulge of a gas-devoid elliptical galaxy, matching the models of \citet{nasim2020}.
Each galaxy is thus represented as a stellar-only Dehnen sphere \citep{dehnen1993} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed with zero initial velocity.
We test five different mass resolutions by sampling the stellar phase space distribution with a varying number of stellar particles: $N_\star = \{2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6\}$.

We then construct isolated merger initial conditions by placing two galaxies on an elliptical orbit with varying values of eccentricity, at a fixed initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
The chosen eccentricities $e_0$ are 0.90 and 0.99.
The radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris2022} for the $e_0=0.90$ orbit.
For each orbital configuration, we run ten realisations for each mass resolution, to account for stochasticity caused by the discretised phase space.

@@

\begin{figure}
    \includegraphics{convergence}
    \caption{Convergence of the eccentricity scatter $\sigma_e$ as a function of mass resolution $M_\bullet/m_\star$ and the number of stellar particles per progenitor galaxy $N_\star$ for initial eccentricities $e_0=0.90$ and $e_0=0.99$. 
    The expected $1/\sqrt{N_\star}$ scaling is recovered for the $e_0 = 0.90$ mergers after applying a $2.5\sigma$-cut to the simulation data. 
    For the $e_0=0.99$ mergers, a steeper scaling than the $1/\sqrt{N_\star}$ is obtained after applying a similar $2.5\sigma$-cut to the data.}
    \label{fig:convergence}
\end{figure}

\subsection{Eccentricity scatter in simulations}

@@

Critically, the eccentricity values beyond $2.5\sigma$ from the mean that are removed in our cleaned samples are each still a valid measurement from the simulation, and highlight the presence of stochasticity in eccentricity for these idealised simulation set-ups with a highly-radial initial orbit.

@@

To understand the observed scatter in eccentricity, we investigate the binary binding process in each simulation. 
Whilst the two SMBHs are unbound, they undergo a number of hyperbolic orbits influenced by the galactic merger potential.
At each pericentre passage, the two SMBHs scatter off each other with some deflection angle $\theta_\mathrm{defl}$, defined:
\begin{equation}\label{eq:theta}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, and $L$ is the magnitude of the SMBH 
binary angular momentum vector and $E$ the binary orbital energy at the time of the pericentre passage \citep{binney2008}.
The deflection angle is a function of the impact parameter $b$ of the SMBH trajectory, however $\theta_\mathrm{defl}$ is able to be more robustly measured than $b$ from the simulation data.
\drafting{
    Each pericentre passage is associated with a deflection angle, however we find that it is the first extreme deflection angle that is important.
    To quantify extreme, we set as $\theta_\mathrm{defl}$ the first 
angle that exceeds an orbit-dependent threshold:
$\theta_{\mathrm{defl,min}}=90\degr$ for the $e_0=0.99$ mergers and $\theta_{\mathrm{defl,min}}=45\degr$ for the $e_0=0.90$ mergers.
}
In general, this corresponds to the second or third actual pericentre passage of the SMBHs in the simulations.
An example of $\theta_\mathrm{defl}$ is shown for two realisations of the $e_0=0.90$ ($e_0=0.99$) orbit in the left (right) panel of \autoref{fig:orbit}, which occurs between the points A and B in the SMBH binary orbit in each inset panel. 
We observe evidence for a relationship between the deflection angle $\theta_\mathrm{defl}$ and the resulting eccentricity at the time the SMBH binary becomes hard, shown in \autoref{fig:theta_e}.

@@

\subsection{Reproducing the eccentricity behaviour with a toy model}\label{ssec:toy}

@@

The resulting model curves are shown together with the simulation data in \autoref{fig:theta_e}.
To correctly match the data in the $e_0=0.90$ case, the model curve has been shifted to the left by $12\degr$,
while in the $e_0=0.99$ case a smaller shift of $6\degr$ is used.
These shifts are likely required due to the rotation of the stellar component tilting the background potential
relative to the SMBH trajectories in the simulations, with the effect being stronger in the $e_0=0.90$ case
due to the less radial merger orbit.
In the $e_0=0.99$ case we can also see that the simulation data has significant scatter around
the model curve in the $\theta_\mathrm{defl} \gtrsim 120\degr$ region,
although the model does appear to capture the mean behaviour fairly well.
The behaviour of the model in this region is also quite sensitive to the exact value of $e_\mathrm{s}$,
which might be related to the large scatter in the simulation data.

Otherwise the behaviour of the simulation data is captured well,
with the model correctly producing the two main eccentricity minima,
as well as the $e\approx 1$ region between them.
The minima occur for trajectories where the torque from the background potential together with the dynamical friction
causes the SMBHs to loop around into a nearly circular orbit,
while highly eccentric binaries are produced when the binary is trapped into nearly radial oscillations along 
the main axes of the potential, as is seen in \autoref{fig:model_orbits}.

@@

Our results show that the variation in eccentricity $e_\mathrm{h}$ for a given system configuration is tightly correlated with the variation in the deflection angle $\theta_\mathrm{defl}$, independent of the simulation mass resolution.
Together with the simple toy model that reproduces well the behaviour seen in the simulations,
this demonstrates that the sensitivity of the SMBH binary eccentricity to slight changes in the merger orbit is a physical feature of galactic mergers on highly radial orbits,
at least in major mergers where the SMBHs have nearly equal masses,
and not an error due to insufficient numerical resolution.

As shown by both the simulations and the toy model, in these mergers the eccentricity of the SMBH binary can span nearly the full possible range between 0 and 1,
depending on comparatively tiny differences on the scale of a parsec in the particular realisation of the galactic merger orbit.
Compared to major mergers, in minor mergers where there is less asymmetry in the stellar background when the SMBHs become bound,
the scatter in eccentricity is likely to be low even in the case of a radial merger orbit, since the system is less sensitive to the exact value of the deflection angle,
or the impact parameter, as can be seen from the $e_\mathrm{s}=0.2$ case in \autoref{fig:model_orbits}.

The simulation setup used in this work is highly idealised,
and covers only a small part of the parameter space of merger configurations since we only consider fairly high-eccentricity galactic merger orbits.
However, similar behaviour can be expected to occur also in more realistic models,
since the relevant dynamics occurs within a few hundred parsecs from the centre of the merged galaxy,
where we do not expect any significant  effects from dark matter or the outer parts of the galaxy, which are not included in our simulations.
Lower eccentricity merger orbits can be expected to show less scatter in the binary eccentricity due to the lack of a hard scattering event that is sensitive to slight perturbations in the merger orbit,
and this is also seen in simulation results \citep{gualandris2022}.
However, most major mergers are expected to occur on highly radial orbits based on cosmological simulations \citep[e.g.][]{khochfar2006},
and the dynamical friction from the dark matter halo would generally drive even lower eccentricity merger orbits to highly radial final plunges when the initial separation of the galaxies is large enough.

In numerical simulations with a discretised phase space, the two primary causes of random variation in the impact parameter are the initial particle sampling variation in the merger initial conditions and the Brownian motion of the SMBH binary \citep{merritt2001}.
Both of these effects scale as $1/\sqrt{N_\star}$, which can explain the observed scaling of $\sigma_e$ \citep{nasim2020} when the system falls into the region of parameter space where the relation between $e$ and $b$
is approximately linear.
The deviation from this scaling seen in \autoref{fig:convergence}, particularly when not excluding outlying data points, is then simply explained by the fact that the eccentricity is not globally a linear function of the impact parameter.

While the scatter in the impact parameter between different numerical realisations of the same merger is due to relatively low number of particles compared to real galaxies,
we expect that uncertainty of a similar magnitude is also present in real systems due various mechanisms, such as perturbations from their environment.
The SMBHs are likely also not located exactly at the centre of mass of the galaxy,
as e.g.\ work by \citet{batcheldor2010} has suggested that the SMBH in M87 may be displaced from the galactic photo-centre by $\sim7\,\mathrm{pc}$ due to jet acceleration or gravitational recoil kicks. 
It is therefore evident that it is not possible to give an exact prediction for the SMBH binary eccentricity produced by a given merger,
and instead we must focus on predicting the distribution of eccentricities that can be produced by a given merger configuration.
Such predictions are crucial for constraining observational quantities that are correlated with binary eccentricity,
most notably SMBH binary merger timescale and the spectrum of GWs at nanohertz frequencies observable with PTAs \drafting{(refs)}.

@@

A.R. acknowledges the support by the University of Helsinki Research Foundation.
P.H.J. acknowledges the support
by the European Research Council via ERC Consolidator Grant KETJU (no. 818930).
